Towards Practical First-Order Model Counting

### 2024-04-23 doc/paper.tex
The rest of Section 4

\section{Preliminaries (TODO)}

\Cref{sec:logic} introduces the fundamental ideas of first-order logic and
WFOMC, and discusses the different logics used in the latter.\@
\Cref{sec:algebra} introduces the terminology we use to describe the output of
the original \textsc{Crane} algorithm~\cite{DBLP:conf/kr/DilkasB23}, i.e.,
functions and equations that define them.

@@

\begin{example}\label{example:overall}
  Let us consider the following formula (previously examined by Dilkas and
  Belle~\shortcite{DBLP:conf/kr/DilkasB23}) that defines predicate $P$ to be a
  bijection between two sets $\Gamma$ and $\Delta$:
  \[
    \begin{gathered}
      \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
      \forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y)\\
      \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z\\
      \forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z.
    \end{gathered}
  \]
  In particular, we examine the first solution that \Cranebfs{} returns for this
  formula.

@@

In the rest of this section, we extend the smoothing algorithm of Van den Broeck
et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11} even further.
\Cref{sec:smoothingbase} describes the role smoothing plays in the
base-case-finding algorithm from \cref{sec:main}. Then,
\cref{sec:smoothingfcg1,sec:smoothingfcg2,sec:smoothingfcg3} show how to adapt
smoothing to the compilation rules introduced by Dilkas and
Belle~\shortcite{DBLP:conf/kr/DilkasB23}.

@@

\subsection{Smoothing the FCG (TODO)}\label{sec:smoothingfcg1}

\todo[inline,caption={}]{
  \begin{itemize}
    \item introduce the relevant node types and their notation
    \item Describe \cref{fig:smoothing1,fig:smoothing2}
  \end{itemize}
}

\begin{figure}
  \centering
  \begin{forest}
    for tree={sn edges}
    [$\DR(c \in \Delta)$,name=dr,ellipse,draw,label={right:\textcolor{blue}{3. $\{\, \forall x \in \Delta\text{. }P(x) \,\}$}}
    [$\land$,ellipse,draw,label={left:\textcolor{blue}{2. $\{\, P(c) \,\}$}},label={right:\textcolor{blue}{6. $\{\, P(c), \forall x \in \Delta\text{. }x \ne c \Rightarrow P(x) \,\}$}}
    [$P(c)$,rectangle,draw,fill=green!20,label={272:\textcolor{blue}{1. $\{\, P(c) \,\}$}}]
    [$\CR(\Delta' \gets \Delta \setminus \{\, c \,\})$,ellipse,draw,label={87:\textcolor{blue}{5. $\{\, \forall x \in \Delta\text{. }x \ne c \Rightarrow P(x) \,\}$}}
    [$\Reff(\Delta \mapsto \Delta')$,name=ref,ellipse,draw,label={below:\textcolor{blue}{4. $\{\, \forall x \in \Delta'\text{. }P(x) \,\}$}}]
    ]
    ]
    ]
  \end{forest}
  \caption{An example based on bijection counting (but simplified a lot). The
    arc from the $\Reff$ node to the $\DR$ node is
    omitted.}\label{fig:smoothing1}
\end{figure}

\begin{figure}
  \centering
  \begin{forest}
    for tree={sn edges}
    [$\DR(c \in \Delta)$,name=dr,ellipse,draw
    [$\land$,ellipse,draw,dashed,edge=dashed
    [$P(c)$,rectangle,draw,fill=green!20]
    [$\forall x \in \Delta\text{. }x \ne c \Rightarrow P(x) \lor \neg P(x)$,rectangle,dashed,edge=dashed,draw,fill=blue!20]
    ]
    ]
  \end{forest}
  \caption{An artificial example of a situation when smoothing nodes need to be
    added below a $\DR$ node. The dashed part is added during
    smoothing}\label{fig:smoothing2}
\end{figure}

Smoothing is a two-step process. First, atoms that are still accounted for in
the circuit are propagated upwards. Then, at nodes of certain types, missing
atoms are detected and additional sinks are created to account for them. If left
unchanged, the first step of this process would result in an infinite loop
whenever a cycle is encountered.

\todo[inline]{Don't say `upwards'}

\subsection{Stage 1: Propagating Unit Clauses (TODO)}\label{sec:smoothingfcg2}

\begin{description}
  \item[\texttt{Ref}.] During smoothing, when unit clauses are propagated in the
        opposite direction of the FCG arcs (i.e., `upwards'), when visiting a
        \texttt{Ref} node, these clauses are translated using the domain map of
        the \texttt{Ref} node. For example, if the domain map include
        $\Delta \mapsto \Delta'$, and the unit clause mentions $\Delta$, then
        replace it by $\Delta'$.
  \item[Constraint removal.] Do reverse constraint removal. Assuming that
        domain $\Delta$ with constraints $x \ne c$ (for some constant
        $c \in \Delta$) were replaced with domain $\Delta'$, replace each
        $\forall x \in \Delta'\text{. }\phi(x)$ with
        $\forall x \in \Delta\text{. }X \ne c \Rightarrow \phi(x)$.
  \item[Domain recursion.] Suppose the domain recursion node introduces constant
        $c \in \Delta$. For each unit clause received from the child node,
        replace each occurrence of $\phi(c)$ or $\forall x \in \Delta\text{.
        } x \ne c \Rightarrow \phi(x)$ with $\forall x \in \Delta\text{.
        }\phi(x)$. This can be seen as a claim about what ground atoms the
        domain recursion node \emph{should} cover (or a temporary assumption).
        If the relevant subgraph indeed covers those ground atoms, Stage~2 will
        do nothing. Otherwise, smoothing nodes will be added below the domain
        recursion node to cover the difference between what was propagated from
        the domain recursion node and what was received from the child node.
\end{description}

\subsection{Stage 2: Adding Smoothing Nodes (TODO)}\label{sec:smoothingfcg3}

We never need to add smoothing nodes after \texttt{Ref} or constraint removal
nodes. However, for domain recursion we must do the following.

\begin{enumerate}
  \item Whenever the set of unit clauses of the child node contains two formulas
        $\phi(c)$ and $\forall x \in \Delta\text{.
        } x \ne c \Rightarrow \phi(x)$ (i.e., the only difference between the
        two formulas is that one has the constant $c$ whereas the other one has
        a variable $x \ne c$), merge them into $\forall x \in \Delta\text{.
        }\phi(x)$.
  \item Add smoothing nodes below the domain recursion node for the difference
        between the unit clauses assigned to the domain recursion node during
        Stage 1 and the unit clauses of the child node post-processed by the
        step above. For example, if the child node `covers' only $P(c)$, then
        Stage 1 assigns $\forall x \in \Delta\text{. }P(x)$ to the domain
        recursion node. The smoothing node below the domain recursion node then
        has the clause $\forall x \in \Delta\text{. } x \ne c \Rightarrow P(x)$.
\end{enumerate}

\todo[inline]{Can I formally define what is meant by `difference'?}

>>>>

\section{Preliminaries}

\Cref{sec:logic} introduces the fundamental ideas of first-order logic and
WFOMC, and discusses the different logics used in the latter.\@ Then,
\cref{sec:algebra} introduces the terminology we use to describe the output of
the original \textsc{Crane} algorithm~\cite{DBLP:conf/kr/DilkasB23}, i.e.,
functions and equations that define them.

@@

\begin{example}\label{example:overall}
  Consider the following formula (previously examined by Dilkas and
  Belle~\shortcite{DBLP:conf/kr/DilkasB23}) that defines predicate $P$ to be a
  bijection between two sets $\Gamma$ and $\Delta$:
  \[
    \begin{gathered}
      \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
      \forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y)\\
      \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z\\
      \forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z.
    \end{gathered}
  \]
  In particular, we examine the first solution that \Cranebfs{} returns for this
  formula.

@@

In the rest of this section, we extend the smoothing algorithm of Van den Broeck
et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11} even further.
\Cref{sec:smoothingbase} describes the role smoothing plays in the
base-case-finding algorithm from \cref{sec:main}. Then, \cref{sec:smoothingfcg1}
shows how to adapt smoothing to the compilation rules introduced by Dilkas and
Belle~\shortcite{DBLP:conf/kr/DilkasB23}.

@@

\subsection{Smoothing the FCG}\label{sec:smoothingfcg1}

Smoothing is a two-step process. First, positive unit clauses (denoting sets of
ground atoms that are accounted for in the FCG) are propagated `upwards', i.e.,
in the opposite direction of FCG arcs. Then, at nodes of certain types, missing
atoms are detected and additional nodes are added to account for them. In this
section, we:
\begin{enumerate*}[label=(\roman*)]
  \item describe the relevant node types from previous work,
  \item show how smoothing ought to work for these node types, and
  \item illustrate how the new smoothing techniques work on two example FCGs.
\end{enumerate*}

Before describing the proposed changes to smoothing, we briefly review the
compilation rules (and their corresponding node types) introduced by Dilkas and
Belle~\shortcite{DBLP:conf/kr/DilkasB23} while referring to the aforementioned
paper for precise definitions. \emph{Domain recursion} selects a domain
$\Delta$, introduces a new constant $c \in \Delta$, and, for each variable $x$
quantified over $\Delta$, considers two possibilities: $x = c$ or $x \ne c$,
modifying the formula accordingly. We denote the resulting node as
$\DR(c \in \Delta)$. \emph{Constraint removal} applies to formulas such that,
for some domain $\Delta$, each variable $x$ quantified over $\Delta$ is followed
by the inequality constraint $x \ne c$. In other words, the relevant clauses can
be rewritten to begin with $\forall x \in \Delta\text{.
}x \ne c \Rightarrow \ldots$ For such formulas, constraint removal replaces
$\Delta$ with a new domain $\Delta'$ and removes the $x \ne c$ constraints. We
denote the resulting node as $\CR(\Delta' \gets \Delta \setminus \{\,c\,\})$.
Finally, \emph{caching} detects when the input formula $\phi$ is equal to a
previously-encountered formula $\psi$ except for having different domains. We
denote the resulting node as $\Reff(\sigma)$, where $\sigma$ is the substitution
mapping the domains of $\psi$ to their corresponding domains in $\phi$.

\paragraph{Stage~1 for domain recursion.}
When visiting a $\DR(c \in \Delta)$ node, for each unit clause received from the
child node, we replace each occurrence of $\phi(c)$ or
$\forall x \in \Delta\text{. } x \ne c \Rightarrow \phi(x)$ with
$\forall x \in \Delta\text{. }\phi(x)$. In other words, if the subgraph below
the domain recursion node covers some of the ground atoms that are affected by
domain recursion, then it should cover all of them. If the relevant subgraph
indeed covers those ground atoms, Stage~2 will do nothing. Otherwise, smoothing
nodes will be added below the domain recursion node to cover the difference
between the sets of ground atoms assigned to the domain recursion node and its
child node.

\paragraph{Stage~1 for constraint removal and caching.}
When visiting a $\CR(\Delta' \gets \Delta \setminus \{\,c\,\})$ node, we reverse
constraint removal. In other words, we replace each
$\forall x \in \Delta'\text{. }\phi(x)$ with $\forall x \in \Delta\text{.
}x \ne c \Rightarrow \phi(x)$. When visiting a $\Reff(\sigma)$ node, we apply
substitution $\sigma$ to the unit clauses coming from the child node.

\paragraph{Stage~2 for domain recursion.}
We need not add smoothing nodes immediately below constraint removal or caching
nodes. However, for domain recursion we do the following.
\begin{enumerate}
  \item Suppose the set of unit clauses assigned to the child node during
        Stage~1 contains both $\phi(c)$ and $\forall x \in \Delta\text{.
        } x \ne c \Rightarrow \phi(x)$. In other words, the only difference
        between the two clauses is that one has the constant $c$ whereas the
        other one has a variable $x \ne c$. In such a case, we merge the two
        clauses into $\forall x \in \Delta\text{. }\phi(x)$.
  \item If necessary, we add smoothing nodes below the domain recursion node to
        cover the difference between the unit clauses assigned to the domain
        recursion node during Stage~1 and the unit clauses of the child node
        post-processed by the step above.
\end{enumerate}

\begin{figure}
  \centering
  \begin{forest}
    for tree={sn edges}
    [$\DR(c \in \Delta)$,name=dr,ellipse,draw
    [$\land$,ellipse,draw,dashed,edge=dashed
    [$P(c)$,rectangle,draw,fill=green!20]
    [$\forall x \in \Delta\text{. }x \ne c \Rightarrow P(x) \lor \neg P(x)$,rectangle,dashed,edge=dashed,draw,fill=blue!20]
    ]
    ]
  \end{forest}
  \caption{An artificial example of an FCG where a smoothing node must be added
    below a domain recursion node. The dashed nodes and arcs are added during
    Stage~2 of smoothing.}\label{fig:smoothing2}
\end{figure}

\begin{example}[An FCG that needs smoothing]
  \Cref{fig:smoothing2} shows an FCG with a domain recursion node
  $\DR(c \in \Delta)$ immediately followed by a single $P(c)$ node. In this
  case, Stage~1 of smoothing assigns $\{\, \forall x \in \Delta\text{.
  }P(x) \,\}$ to the former node and $\{\, P(c) \,\}$ to the latter. Since these
  two sets of unit clauses cover different ground atoms, Stage~2 adds a
  smoothing node to cover $P(x)$ for all $x \in \Delta \setminus \{\, c \,\}$.
\end{example}

\begin{figure}
  \centering
  \begin{forest}
    for tree={sn edges}
    [$\DR(c \in \Delta)$,name=dr,ellipse,draw,label={right:\textcolor{blue}{3. $\{\, \forall x \in \Delta\text{. }P(x) \,\}$}}
    [$\land$,ellipse,draw,label={left:\textcolor{blue}{2. $\{\, P(c) \,\}$}},label={right:\textcolor{blue}{6. $\{\, P(c), \forall x \in \Delta\text{. }x \ne c \Rightarrow P(x) \,\}$}}
    [$P(c)$,rectangle,draw,fill=green!20,label={272:\textcolor{blue}{1. $\{\, P(c) \,\}$}}]
    [$\CR(\Delta' \gets \Delta \setminus \{\, c \,\})$,ellipse,draw,label={87:\textcolor{blue}{5. $\{\, \forall x \in \Delta\text{. }x \ne c \Rightarrow P(x) \,\}$}}
    [$\Reff(\Delta \mapsto \Delta')$,name=ref,ellipse,draw,label={below:\textcolor{blue}{4. $\{\, \forall x \in \Delta'\text{. }P(x) \,\}$}}]
    ]
    ]
    ]
  \end{forest}
  \caption{A smooth FCG loosely based on \cref{example:overall}. The arc from
    the caching node to the domain recursion node is omitted for compactness.
    The labels next to nodes show the sets of unit clauses assigned to each node
    during Stage~1 and the order of these assignments, omitting empty
    assignments that replace a set $S$ with $S$ itself.}\label{fig:smoothing1}
\end{figure}

\begin{example}[A smooth FCG]\label{example:smooth}
  Stage~1 of smoothing is more involved in the case of the FCG in
  \cref{fig:smoothing1}. The unit clause $P(c)$ propagates to the conjunction
  node and is then generalised to $\forall x \in \Delta\text{. }P(x)$ by the
  domain recursion node. It then propagates to the caching node, changing form
  to $(\forall x \in \Delta\text{.
  }P(x))[\Delta \mapsto \Delta'] \equiv \forall x \in \Delta'\text{. }P(x)$. The
  constraint removal node re-introduces the constraints, transforming the clause
  to $\forall x \in \Delta\text{. }x \ne c \Rightarrow P(x)$, which is then
  propagated to the conjunction node, joining $P(c)$.

  In Stage~2, $P(c)$ and $\forall x \in \Delta\text{. }x \ne c \Rightarrow P(x)$
  are merged into $\forall x \in \Delta\text{. }P(x)$. Since the resulting
  clause matches the clause assigned to the domain recursion node, the FCG is
  already smooth.
\end{example}

The algebraic interpretation of \cref{fig:smoothing1} is an equation $e$ that
defines a recursive function. The right-hand side of $e$ already covers $P(c)$,
and the smoothing algorithm correctly recognises that the recursive call also
covers $P(x)$ for all $x \in \Delta \setminus \{\, c \,\}$.
